### 2021-05-17 math/ia_2021/paper.tex
ia 2021 final....

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
        \vspace*{4cm}
        A statistical approach to predicting a tsunami’s wave height \\
        \vspace{1cm}
        Can the height of a tsunami's wave be predicted based on the distance from and
        the magnitude of a preceding earthquake? \\
        \vspace{1cm}
        \textit{Väinö-verneri Kauppila} \\
        February 2021 \\
        \vspace{4cm}
        Word count: TBD \\
        \textit{First draft, details are missing}
        \vfill
        \vspace{0.1cm}
    \end{center}
\end{titlepage}

@@

The dataset, procured from the National Oceanic and Atmospheric Administration website \cite{noaa}, includes many tsunami runups, going back before the year 0. Due to this, the NOAA states uncertainties that can occur in its data. For example, reports that were recorded before the 20th century tend to be based on written accounts, and not recorded seismic activity. Only since the 1960s, the installation of the Worldwide Standardized Seismograph Network system allowed for more accurate data in this paper. For that reason, any recorded activity before this date was discarded from the dataset for this paper. The dataset itself can be accessed through two different methods; searching the database using the online tool or downloading it in a somewhat obscure compressed Google Earth \verb|.kmz| format. The online tool supports the download of search results in a much more human-readable \verb|.tsv| (tab-separated) format. For this paper, all entries after the year 1900 were searched and downloaded.

@@

$y =  \beta_0 a + \beta_1 x + e $,

where the term $\beta_0$ represents the y-intercept of the line and $\beta$ represents the coefficient of $x$. The final term $e$ is what is known as the error or residual term. It is an estimate of the amount by which the predictions differ from the real value. An example of linear regression can be found in Appendix \ref{app:regression}, where a set of data points is presented on a scatter plot, and the regression line is found by finding the individual coefficients $\beta_n$ of $x$ (see equation above), like such:

@@

So the relationship between the two independent variables and the wave height can be estimated with a multivariate, polynomial regression model. We will use the Python scikit-learn library to aid in creating this model. \\

(\textit{Section not done, need to do more research on how the} \verb|PolynomialFeatures| \textit{function transforms a linear array to one fit for polynomial regression})

@@

@@

The research included in this paper could be useful for many life-saving scenarios, notably ones in which, for example, high quality and large-scale tsunami alert systems are not available.

\textit{Conclusion and evaluation of the method are contradictory, to be corrected}

@@

It has also come to my attention that using the $R^2$ value for a non-linear relationship is not suitable, as it does not follow the same characteristics.

@@

Below is the dataset plotted, along with a line of regression.

>>>>

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
        \vspace*{4cm}
        A statistical approach to predicting a tsunami’s wave height \\
        \vspace{1cm}
        Can the height of a tsunami's wave be predicted based on the distance from and
        the magnitude of a preceding earthquake? \\
        \vspace{1cm}
        \textit{Väinö-verneri Kauppila} \\
        February 2021 \\
        \vspace{4cm}
        \input{wordcount.txt} \\
        \vfill
        \vspace{0.1cm}
    \end{center}
\end{titlepage}

@@

The dataset, procured from the National Oceanic and Atmospheric Administration website \cite{noaa}, includes many tsunami runups, going back to the year 2000 BC. Due to this, the NOAA states uncertainties that can occur in its data. For example, reports that were recorded before the 20th century tend to be based on written accounts, and not recorded seismic activity. Only since the 1960s, the installation of the Worldwide Standardized Seismograph Network system allowed for more accurate data in this paper. For that reason, any recorded activity before this date was discarded from the dataset for this paper. The dataset itself can be accessed through two different methods; searching the database using the online tool or downloading it in a somewhat obscure compressed Google Earth \verb|.kmz| format. The online tool supports the download of search results in a much more human-readable \verb|.tsv| (tab-separated) format. For this paper, all entries after the year 1900 were searched and downloaded.

@@

$y =  \beta_0 x + \beta_1 x + e $,

where the term $\beta_0$ represents the y-intercept of the line and $\beta$ represents the coefficient of $x$. The final term $e$ is what is known as the error or residual term. It is an estimate of the amount by which the predictions differ from the real value. An example of linear regression can be found in Appendix \ref{app:regression}. The set of data points is presented here on a scatter plot (along with its line of regression, explained in the following paragraph):

The regression line is found by finding the individual coefficients $\beta_n$ of $x$ (see equation above), like such:

@@

So the relationship between the two independent variables and the wave height can be estimated with a multivariate, polynomial regression model. The Python scikit-learn library was used to aid in creating this model. The model itself can be found in Appendix \ref{app:scripts}.

@@

This function is implemented in the sklearn library, and when used on our model, we get the result of $R^2 = 0.36$ (rounded to 2 s.f.).

This result is not optimal, as it means that only $\frac{1}{3}$ of the data from our model fits the test dataset. However, the $R^2$ value does not explain anything about causation, which is obviously present in this data, as it could be said that logically the distance from a wave is proportional to its observed height.

@@

The research included in this paper would not be very useful for life-saving scenarios, because of its low $R^2$ value. However, further research by professionals on this subject could potentially reveal relationships between these variables and as such create better models. A low score such as the one in this paper can be due to many things, but mainly I think this was due to the fit of the model. An exponential regression model would fit better.

@@

It has also come to my attention that using the $R^2$ value for a non-linear relationship is not suitable, as it assumes the relationship is linear \cite{frost_2018}.

@@
